On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-09-02 paper.tex
Added note
Alex proofs of maximum and expected tiles touched
Added proofs the tiles-touched formula (Theorem 3) and expected tiles visited via Buffon's Needle argument (Theorem 6).

The notion of a \emph{random} segment of a given length is precisely defined as follows. By symmetry, one endpoint of the segment can be assumed to lie in a fixed, reference tile. The position of this endpoint is \emph{uniformly} distributed on the tile. The segment orientation has a \emph{uniform} distribution on the set of all possible directions, and is \emph{independent} of the endpoint position. Solving the problem of how many tiles the segment visits on average, as will be seen, also answers the inverse question (segment length to visit a given number of tiles on average). A natural, related question is with that probability the segment visits the maximum number of tiles.

@@

@@

\begin{theorem}
\label{theo: funl, form}
For $a, b \in \mathbb R^+$ and $\tiles \in \mathbb N$,
\begin{equation}
\label{eq: theo: funl, form; funl}
\funl(\tiles) = \sqrt{(\isoli-2)^2 a^2 + (\jsoli-2)^2 b^2}
\end{equation}
with
\begin{align}
\label{eq: theo: funl, form; i}
\isoli &= \max\left\{ \left\lfloor \frac {(2\tiles-1)b^2+5a^2}{2(a^2+b^2)} \right\rfloor, 2 \right\}, \\
\label{eq: theo: funl, form; j}
\jsoli &= \max\left\{ \left\lceil \frac {2\tiles a^2 + 3(b^2-a^2)}{2(a^2+b^2)} \right\rceil, 2 \right\}.
\end{align}
Equivalently,
\begin{equation}
\text{***new formula found by Alex; unless we can unify with the above}
\end{equation}
This function is monotone increasing for $\tiles \geq 3$.
\end{theorem}

@@

For $a=b=1$ the set $\mss$ consists of points of the form $(i,i)$ and $(i,i-1)$, as is easily seen from Proposition~\ref{prop: min suff set, form}, and as illustrated in Figure~\ref{fig: pairs_bounds_1}. By symmetry, replacing each point $(i,i-1)$ by $(i-1,i)$ gives a set $\mss'$ which is also minimal sufficient. For this new set, the lower bounding line \eqref{eq: lower bound, line} can be replaced by the simpler $j=i$. The same approach used in the proof of Theorem~\ref{theo: funt, form} can be followed, but using this line. Thus $(\isolr, \jsolr)$ is obtained from
\begin{align}
(\isolr-2)^2 + (\jsolr-2)^2  &= \len^2, \\
\jsolr &= \isoli,
\end{align}
which gives
\begin{equation}
\label{eq: theo: funt, sq, form; proof 1}
\isolr = \len/\sqrt{2} + 2.
\end{equation}
The pair $(\isoli,\jsoli)$ resulting from \eqref{eq: theo: funt, sq, form; proof 1} is in $\mss'$, and is given in \eqref{eq: theo: funt, sq, form; i} and \eqref{eq: theo: funt, sq, form; j}. This establishes \eqref{eq: theo: funt, sq, form; funt, with i, j}.

***Alex
***Derivation of \eqref{eq: theo: funt, sq, form; funt, simpler} \ldots
***It looks like we may have to use different techniques for deriving \eqref{eq: theo: funt, sq, form; funt, with i, j} and \eqref{eq: theo: funt, sq, form; funt, simpler}. I don't see a way to use a common proof method. And \eqref{eq: theo: funt, sq, form; funt, with i, j} is needed because of $\isoli$, $\jsoli$, which are used in (the text below in preparation for) \S\ref{part: probmax: unit square grid, real lengths}.
\end{proof}

@@

Given $\len \in \mathbb R^+$, consider a segment of length $\len$ with random position and orientation. Specifically, the coordinates $x_1$, $y_1$ of the first endpoint are independent random variables uniformly distributed on $[0,a)$ and $[0,b)$ respectively. The orientation $\theta$ of the segment is uniformly distributed on $[0,2\pi)$. The variables $x_1$, $y_1$ and $\theta$ determine the coordinates $x_2$, $y_2$ of the second endpoint.

@@

\begin{proof}
***New proof without using the proposition (which has been moved after the theorem)
\end{proof}

>>>>

% *** AA: Maybe add a transition that now we're switching to looking at the average number of tiles visited by a random segment.
The notion of a \emph{random} segment of a given length is precisely defined as follows. By symmetry, one endpoint of the segment can be assumed to lie in a fixed, reference tile. The position of this endpoint is \emph{uniformly} distributed on the tile. The segment orientation has a \emph{uniform} distribution on the set of all possible directions, and is \emph{independent} of the endpoint position. Solving the problem of how many tiles the segment visits on average, as will be seen, also answers the inverse question (segment length to visit a given number of tiles on average). A natural, related question is with that probability the segment visits the maximum number of tiles.

@@

% ***Is the text from here to the proposition too long? Shorten it?

% ***AA: Do we actually need the concept of a minimal sufficient set, given that there's a single value at each diagonal (tiebreaking a certain way)? Perhaps rename to suggest it's one value?

@@

\begin{theorem}
\label{theo: funl, form}
For $a, b \in \mathbb R^+$ and $\tiles \in \mathbb N$,
\begin{equation}
\label{eq: theo: funl, form; funl}
\funl(\tiles) = \sqrt{(\isoli-2)^2 a^2 + (\jsoli-2)^2 b^2}
\end{equation}
with
\begin{align}
\label{eq: theo: funl, form; i}
\isoli &= \max\left\{ \left\lfloor \frac {(2\tiles-1)b^2+5a^2}{2(a^2+b^2)} \right\rfloor, 2 \right\}, \\
\label{eq: theo: funl, form; j}
\jsoli &= \max\left\{ \left\lceil \frac {2\tiles a^2 + 3(b^2-a^2)}{2(a^2+b^2)} \right\rceil, 2 \right\}.
\end{align}
Equivalently,
\begin{equation}
\text{***new formula found by Alex; unless we can unify with the above}
% *** AA: Wrote a shot at how a unification might look in e-mail.
\end{equation}
This function is monotone increasing for $\tiles \geq 3$.
\end{theorem}

@@

For $a=b=1$, the set $\mss$ consists of points of the form $(i,i)$ and $(i,i-1)$, as is easily seen from Proposition~\ref{prop: min suff set, form}, and as illustrated in Figure~\ref{fig: pairs_bounds_1}. By symmetry, replacing each point $(i,i-1)$ by $(i-1,i)$ gives a set $\mss'$ which is also minimal sufficient. For this new set, the lower bounding line \eqref{eq: lower bound, line} can be replaced by the simpler $j=i$. The same approach used in the proof of Theorem~\ref{theo: funt, form} can be followed, but using this line. Thus $(\isolr, \jsolr)$ is obtained from
\begin{align}
(\isolr-2)^2 + (\jsolr-2)^2  &= \len^2, \\
\jsolr &= \isoli,
\end{align}
which gives
\begin{equation}
\label{eq: theo: funt, sq, form; proof 1}
\isolr = \len/\sqrt{2} + 2.
\end{equation}
The pair $(\isoli,\jsoli)$ resulting from \eqref{eq: theo: funt, sq, form; proof 1} is in $\mss'$, and is given in \eqref{eq: theo: funt, sq, form; i} and \eqref{eq: theo: funt, sq, form; j}. This establishes \eqref{eq: theo: funt, sq, form; funt, with i, j}.

***Alex
***Derivation of \eqref{eq: theo: funt, sq, form; funt, simpler} \ldots
***It looks like we may have to use different techniques for deriving \eqref{eq: theo: funt, sq, form; funt, with i, j} and \eqref{eq: theo: funt, sq, form; funt, simpler}. I don't see a way to use a common proof method. And \eqref{eq: theo: funt, sq, form; funt, with i, j} is needed because of $\isoli$, $\jsoli$, which are used in (the text below in preparation for) \S\ref{part: probmax: unit square grid, real lengths}.

To show \eqref{eq: theo: funt, sq, form; funt, simpler}, we refer back to \eqref{eq: theo: funt, sq, form; funl} in Corollary~\ref{theo: funl, sq, form} for $\tiles \geq 3$:
\begin{equation}
\funl(\tiles) = \sqrt{\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil}
\end{equation}
Then, the pseudo-inverse function $\funt(\len)$ is given by the largest positive integer $t$ such that 
\begin{equation}
\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil < \len ^ 2
\end{equation}
Since the left-hand side is an integer, the condition of being strictly less than $\len^2$ is equivalent to: 
\begin{equation}
\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil  \leq \left\lceil \len^2 \right\rceil  - 1 
\end{equation}
and removing the now-superfluous ceiling on the left side gives:
\begin{equation}
\frac{(\tiles-3)^2} {2}\leq  \left\lceil \len^2 \right\rceil  - 1 
\end{equation}
Finally, we solve for $\tiles$:
\begin{equation}
\tiles \leq \sqrt{2 \left\lceil \len^2 \right\rceil -2 } + 3
\end{equation}
The desired quantity $\funt(\len)$, that is the largest positive integer $\tiles$ satisfying the above, is thus the floor of the right-hand side as given by \eqref{eq: theo: funt, sq, form; funt, simpler}.
\end{proof}

@@

Given $\len \in \mathbb R^+$, consider a segment of length $\len$ with uniformly random position and orientation. Specifically, the coordinates $x_1$, $y_1$ of the first endpoint are independent random variables uniformly distributed on $[0,a)$ and $[0,b)$ respectively. The orientation $\theta$ of the segment is uniformly distributed on $[0,2\pi)$. The variables $x_1$, $y_1$ and $\theta$ determine the coordinates $x_2$, $y_2$ of the second endpoint.

@@

\begin{proof}
% *** Is this writeup too informal/non-rigorous?
The classic Buffon's Needle result (*** cite Buffon's 1777 paper?) says that if we take infinite ruled paper with parallel lines a distance of $1$ apart, and we toss a needle of length $1$ with uniformly random position and orientation, the probability that it crosses any line is $2/\pi$.

This is also the expected (that is, average) number of crossings, since this needle cannot reach more than one line\footnote{We may ignore the probability-zero event of it lying exactly along a line.}. For a needle of length $2$, the expected number of crossings is double the amount of $2/\pi$, since expectation is additive and so unchanged if we instead break this needle into two length-$1$ halves
that we toss separately and count their total crossings. Conversely, for a needle of length $1/2$, it would be half the original expectation.

A similar argument extends this proportionality to all rational lengths, and by monotonicity all real lengths: for a needle of length $l$, the expected number of crossings is
$\frac{2\len}{\pi}$.
If the lines are a distance $a$ apart, this is equivalent to shrinking the needle by a factor of $a$, giving expected crossings
$\frac{2\len}{\pi a}$.

Now, consider a rectangular grid paper with parameters of width $a$ and height $b$. Equivalently, this is paper ruled with horizontal lines $b$ apart and vertical lines $a$ apart. The distribution of the number of crossings of a needle tossed onto this grid is known as the Buffon-Laplace Needle Problem (*** cite). This number decomposes into the counts of crossings of horizontal and vertical grid lines, so the expected number of crossings is the sum of those for lined paper of width $a$ and $b$ respectively:
\begin{equation}
\frac{2\len}{\pi}\left(\frac \len a + \frac \len b\right)
\end{equation}
As noted in Proposition \ref{prop: i+j-1}, the number of grid cells visited by a segment is the count of its grid line crossings plus one -- unless, that is, it exactly crosses any grid points, but that occurs with probability zero and so doesn't affect the expectation. So, $\funta(\len)$ is the above quantity plus one.

\end{proof}
